Parallel Adaptive Importance Sampling

### 2018-01-25 response_JUQ.tex

\begin{document}
\title{Response to Referee Comments: Parallel Adaptive Importance Sampling}
\maketitle
First of all, we would like to thank the referees and the associate editor for
their time in reading and commenting on our manuscript. In this document
we will address each of the points raised, and describe any relevant
changes that have been made to the paper.

\section*{Referee 1 comments}

\comment{The PAIS algorithm does exhibits interesting ideas to combine the information 
obtained from an ensemble of proposals. Furthermore, it seems quite flexible 
since several proposals and re-sampling methods can be used.

However, the examples somehow fail to convince that the algorithm can be used 
in a wide range of challenging inverse problems. The approach is limited by the 
dimensionality of the problem.}{We thank the referee for their comments. We are very open about the fact that this approach cannot be used for high dimensional problems, but there are a plethora of lower dimensional inverse problems which exhibit structure in the posterior, such as multimodality or high correlations, for which existing MCMC methods struggle to converge in reasonable time. Dimensionality is not the only challenge faced in inverse problems, and these methods have been developed to tackle problems of a different nature. In particular, this method was designed with a view to sampling from inverse problems arising in Biology, where challenging posteriors exist on relatively low dimensional state spaces, and for which the likelihood can be very expensive to compute. We have added some words on this to clarify further the intended type of problem that this method is in the introduction, and again in the discussion at the end.}  %done

\comment{For mixture problems, if the mixture has to many 
elements (or the number of elements is unknown), then this scheme will have 
difficulties.}{This is true, as would state-of-the-art Riemannian manifold MALA, HMC and a whole range of other methods. I would be happy for the referee to indicate a method which would comfortably deal with a mixture problem with a large number of possible known/unknown components. The example was chosen to demonstrate the remarkable performance of the algorithm (with respect to standard MCMC methods) for multimodal posteriors in low dimensions, and we believe it has achieved that. We have added a paragraph to the end of the example to discuss the point that the referee makes.}%done

\comment{Also, it is not clear how it will behave in real life inverse 
problems where the likelihood is determined by complex physical systems that 
involve, for example, ODE's or PDE's. If the likelihood is expensive to 
evaluate, the tunning of the scaling parameters for the PAIS-RW might be a 
concern when the number of chains is big.}{The tuning of the scaling parameters is actually very fast, often converging within a few percentage points of the optimum within the order of $10^2$ iterations. We have included some plots and discussion to the end of Section 4.2 to demonstrate this point, and we thank the referee for highlighting this issue. This is quicker than one might expect for an adaptive serial algorithm.}%done

\comment{An real data example for the PAIS algorithm is much needed. It will be 
interesting to see how PAIS is compared not only against a naive 
parallelization, but also compared against a state of the art MCMC algorithm 
or/and a efficient ad hoc MCMC algorithm designed for such example.}{Unfortunately we do not have the resources to apply our method to a full-scale real data example. However, following additional comments from the editor, we have included a new example where the likelihood requires the solution of an ODE. In Section **** we include an example of data assimilation for the Lorenz*** equations, which is a set of equations which is commonly used to test methods for inference in the applied mathematics community.

Expensive likelihoods actually make this method more applicable, since they make the extra overheads of the resampling step relatively less important. This is precisely what the method is designed for, and is the very reason for us to compare the algorithms by looking at error as a function of the number of samples/likelihood evaluations. We have also added further discussion regarding this to Section ****.}

\comment{There aren't details about the architecture used to run the M chains in 
parallel. This is relevant since there are several options such as: multi-core 
and multi-processor computers, GPU cards, clusters, cloud computing, etc. 
Comments about the advantages and disadvantages of the architecture 
are also welcome. The same goes for details regarding the programming language 
used.}{We thank the referee for this comment, as we should have included these details in the manuscript. The method was implemented in C++, serially, and was run on a single core of a Dell server with four 8-core 3.3GHz CPUs, 64Gb memory, and these details have been added in Section 7.

We are not computer scientists, and as such, we are not in a position to comment on specific architectures, especially as the optimal architecture would be problem specific. The optimal architecture would likely depend on the relative costs of communication of the states across the processors, the cost of the likelihood evaluation, and the cost of the resampler for the chosen ensemble size. We have added a discussion to Section 3, following the presentation of the PAIS algorithm.}

\comment{CPU time must be reported to compare the gain of each algorithm in 
order to have a fair comparison. Authors claim that there is a significant 
gain in terms of the number of iterations needed to achieve a desire 
tolerance; examples in sections 7.2 and 7.3 show that PAIS saves up to 90\% of 
the iterations. However, if the CPU time needed to run 10 PAIS iterations is 
ten times the time needed to run 100 RWMH iterations, then the gain is not as 
good as presented. PAIS iterations are more time consuming than RWMH iterations 
due to the re-sampling step. The authors mentioned that the re-sampling step is 
computationally expensive and that is why they proposed the AMR re-sampling 
method. CPU time is of particular importance in example 7.4 since the number of 
parallel chains grows from 50 to 500.}{The referee may have missed Figure 5.1 in which we show the computational cost, in time, of the resamplers that we use in the paper, both the optimal transport resampler, and the AMR. As mentioned before in response to another of the referee's comments, we had neglected to share the specs of the machine that these times were recorded on, and this has now been added. % There is an additional cost in the algorithm over and above the resampler, which is the computation of the denominator in the weights, as pointed out by the second referee. If we have $M$ members of our ensemble, then the cost of this denominator across the whole ensemble is $M^2$, but it can be easily parallelised. We include some timings of the cost of this aspect of the algorithm in Section ****, and a discussion of the overall additional cost of the approach. 
% Coupled with the timings presented for the resampler, we believe that this now gives the reader a clear idea of the additional overhead costs which are incurred when using our method as opposed to a standard MCMC method. We thank the referee for their comments on this.

The reason for analysing the error as a function of the number of the likelihood evaluations, is that this method is intended for problems with expensive likelihoods, for which these overhead costs would be dwarfed. We have added a discussion on this to the beginning of Section 7.}%done

### 2018-01-25 PAIS.tex
Paper changes + response
Added a reference to the paper as requested. First draft of response
letter.
Referee 1 comments addressed
Apart from the Lorenz example.

Since the clock speed of an individual processor is no longer
following Moore's law\cite{moore1998cramming}, improvements in
computational power are largely coming from the parallelisation of
multiple cores. As such, the area of parallel MCMC methods is becoming
increasingly of interest. One class of parallel MCMC method uses
multiple proposals, with only one of these proposals being
accepted. Examples of this approach include multiple try
MCMC\cite{liu2000multiple} and ensemble MCMC\cite{neal2011mcmc}. In
\cite{calderhead2014general}, a general construction for the
parallelisation of MCMC methods was presented, which demonstrated
speed ups of up to two orders of magnitude when compared with serial
methods on a single core.

@@

In this paper, we present a framework for parallelisation of
importance sampling, which can be built around many of the current
Metropolis-based methodologies in order to create an efficient target
distribution from the current
ensemble. The method makes use of a resampler based on optimal
transport which has been used in the context of particle
filters\cite{reich2013nonparametric}. In particular we demonstrate the
advantages of this method when attempting to sample from multimodal
posterior distributions, such as those arising from inference for
mixture models.

@@

@@

@@

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with ensemble size $M$ for the three different
samplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

@@

\section{Numerical Examples}\label{Sec:Num}

@@

@@

The PAIS has a number of favourable features, for example the
algorithm's ability to redistribute, through the resampling regime,
the ensemble members to regions which require more exploration. This allows the
method to be used to sample from complex multimodal distribution.

@@

>>>>

Since the clock speed of an individual processor is no longer
following Moore's law\cite{moore1998cramming}, improvements in
computational power are largely coming from the parallelisation of
multiple cores. As such, the area of parallel MCMC methods is becoming
increasingly of interest. One class of parallel MCMC method uses
multiple proposals, with only one of these proposals being
accepted. Examples of this approach include multiple try
MCMC\cite{liu2000multiple} and ensemble MCMC\cite{neal2011mcmc}. In
\cite{calderhead2014general}, a general construction for the
parallelisation of MCMC methods was presented, which demonstrated
speed ups of up to two orders of magnitude when compared with serial
methods on a single core. Another approach involves pre-fetching, where the possible future acceptances/rejections are calculated in advance\cite{angelino2014accelerating}.

@@

In this paper, we present a framework for parallelisation of
importance sampling, which can be built around many of the current
Metropolis-based methodologies in order to create an efficient target
distribution from the current
ensemble. The method makes use of a resampler based on optimal
transport which has been used in the context of particle
filters\cite{reich2013nonparametric}. We also detail how algorithmic
parameters can be quickly tuned into optimal regimes, with respect to
the effective sample size statistic. In particular we demonstrate the
advantages of this method when attempting to sample from multimodal
posterior distributions, such as those arising from inference for
mixture models. The method that we present is specifically designed to
tackle the challenges of certain types of low-dimensional inverse
problem which have complex posterior structure, for example
multimodality or strong correlations, and expensive
likelihood evaluations. We will demonstrate that this methodology
allows us to sample from such a posterior distribution with greater
precision per likelihood evaluation than an uncorrelated ensemble of
standard MCMC chains of the same type.

@@

In each iteration of the algorithm, we are required to compute the
likelihood for each member of the ensemble. In the case where the
likelihood is expensive to compute, for example because it requires
the numerical approximation of an ODE or PDE, and/or there is a very
large amount of observations in the data, it will be expedient to
split this computational effort across multiple cores. Whether this is
efficient will depend on the architecture of the machine, and the
relative cost of communication across the cores, to the cost of the
likelihood evaluation. Evaluation of the denominator in the weights
also contributes to the overhead costs of this approach over and above
a standard MCMC method, but this too can be parallelised efficiently,
depending on communication costs of the architecture of the machine.

@@

\begin{figure}[!htb]
\centering
\subfigure[APAIS-RW algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISRW_adapt_var"}}
\subfigure[APAIS-RW algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISRW_adapt_ess"}}
\subfigure[APAIS-MALA algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISMALA_adapt_var"}}
\subfigure[APAIS-MALA algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISMALA_adapt_ess"}}
\caption{Demonstration of the convergence of the adaptive scaling parameter in the adaptively tuned PAIS algorithms.}
\label{fig:G1_adapt_trace_PAIS}
\end{figure}

In the context of an expensive likelihood evaluation, slow convergence
of the algorithmic parameters in this adaptive regime would be a
concern. However, this is rarely an issue, due in part to the speed at
which the PAIS algorithm completes the burn-in phase. Figure
\ref{fig:G1_adapt_trace_PAIS} demonstrates this for two different
versions of the PAIS methodology with a RW proposal and a MALA\cite{roberts1998optimal}
proposal. This picture is typical in our experience, with the scaling
parameters converging in $\mathcal{O}(10^2)$ iterations to an optimal regime.
%%%%%%

@@

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with ensemble size $M$ for the three different
samplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods, where timings have been taken from simulations on a Dell
server with four 8-core 3.3GHz CPUs and 64Gb memory. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

@@

\section{Numerical Examples}\label{Sec:Num}
The numerical examples in this section were all computed on a single
core of Dell
server with four 8-core 3.3GHz CPUs and 64Gb memory. Throughout we
measure the efficiency of the algorithms in terms of the number of
iterations (i.e. likelihood evaluations) which are required in order
to reach a given order of accuracy. The reason for this is that this
method is designed with a particular type of challenging inverse
problem in mind, namely one which is low dimensional, has an expensive
likelihood (such as a PDE solve, and/or very large data) which dwarfs
the overhead cost of PAIS, and which could have complex posterior
structure, such as correlations, multimodality or other highly
non-Gaussian features.

@@

In this example, we have only considered a relatively low dimensional
mixture model problem, with only two distributions in the
mixture. With more elements in the mixture, and/or an undefined number
in the mixture, the dimension of this problem will quickly
increase. Importance sampling schemes such as this suffer from the
curse of dimensionality, limiting the size of the dimension of target
density that it can efficiently sample from. However, this example
does demonstrate the remarkable convergence of the PAIS for multimodal
target distributions.

@@

The PAIS has a number of favourable features, for example the
algorithm's ability to redistribute, through the resampling regime,
the ensemble members to regions which require more exploration. This allows the
method to be used to sample from complex multimodal distributions.

@@

We have demonstrated that the framework that we have considered, with
the use of state-of-the-art optimal transport-based resampling, can
reduce the number of likelihood evaluations required to characterise
complex posterior distributions in low dimensions to a given
degree. We have also introduced a greedy approximation to this
resampler, which drastically reduces the cost, at the loss of some
accuracy, which can be mitigated with the use of larger ensembles.
